Toward GPU-accelerated Traffic Simulation and Its Real-Time Challenge

### 2012-09-16 draft.tex
changed the title
refactored the paper

\title{Accelerating Traffic Simulation and Its Real-Time Challenges}

\author {
Manato Hirabayashi, Shinpei Kato and Masato Edahiro\\
\textit{Department of Information Engineering}\\
\textit{Nagoya University}\\
\and
Yuki Sugiyama\\
\textit{Department of Physics}\\
\textit{Nagoya University}\\
}

@@

\begin{abstract}
 Traffic simulation is a growing domain of computational physics.
 There are many applications in our industry and life that benefit from
 traffic simulation.
 A core challenge of this research domain, however, is posed due to
 an unbounded scale of computation.
 In this paper, we explore an advantage of using the graphics processing
 unit (GPU) to support its computation.
 Two schemes are presented to study how to maximize GPU
 performance in this context.
 Our evaluation using a commodity graphics card shows that our GPU
 implementation improves simulation time by about five times over the
 original CPU implementation.
 We also demonstrate by experiments that additional orders-of-magnitude
 improvements could be achieved by overcoming the current hardware
 limitation.
\end{abstract}

@@

Transportation systems underlie our industry and life as part of
societal infrastructure.
Economy depends highly on traffic flow.
Once a traffic jam occurs, for example, it could cause significant
economic damage through degradation of transport efficiency, energy
consumption, and environmental poisoning.
According to the Japanese government report a few years ago, economic
losses due to traffic congestion reach one hundred billion dollars per
year in Japan.
Albeit a major issue of transportation systems, the mechanism of traffic
jam is not well-explained in the literature.
What is particularly important is ``phantom jam'', which often happens
on freeways without any accident.
This phantom jam should be avoided by science, given that it is
a physical phenomenon but not one caused by human errors.
However, we are not aware of scientific and engineering solutions to
such real-world problems.

In physics, traffic flow is described by mathematical
models~\cite{Bando1995, Kerner1993, Nagel1992}.
While their mathematical expressions are not identical, they are all
compute-intensive procedures.
For example, traffic simulation based on the Optimal Velocity (OV)
model~\cite{Bando1995} computes locations and velocities of agents every
sampling period, solving OV equations. 
Specifically, the location $x_n$ of the $n$th agent is described by the
following equation, where $\Delta x_n = x_{n+1} - x_n$ is a distance to
a preceding agent, $a$ is a sensitivity, and $V()$ is an optimal
velocity function:
\begin{eqnarray}
 \label{eqn:ov}
 \frac{d^2 x_n}{d t^2} = a \left\{V(\Delta x_n) - \frac{d x_n}{d t}\right\}.
\end{eqnarray}

@@

Deployment of traffic simulation may require real-time feedback from the
real world.
This turns out to be a cyber-physical systems (CPS) problem.
Real-time traffic simulation is another challenging issue, where
the rate and the preciseness of simulation must be traded to meet the
requirement of a given scenario.
For instance, evacuation may want very high rate simulation, sacrificing
the preciseness of simulation to some extent.
Unfortunately, none of those challenges has been explored yet, largely
due to a lack of multidisciplinary research collaborations between physics
and computer science.

@@

The rest of this paper is organized as follows.
Section~\ref{sec:assumption} describes the assumption and terminology
used this paper.
Section~\ref{sec:traffic_simulation} briefly explains our traffic
simulation program.
Section~\ref{sec:gpu_implementations} presents our schemes of GPU
implementations of traffic simulation.
Section~\ref{sec:evaluation} evaluates the benefit of our GPU
implementations, and provides an insight into future work.
Section~\ref{sec:conclusion} concludes this paper.

@@

@@

The motivation to study traffic flow is to understand the property of
traffic flow, mainly why traffic congestion occurs.
The study of traffic flow has a long history, and many engineering and
physical models exist.
This paper is particularly focused on the OV model~\cite{Bando1995}.
In this model, a driver is supposed to maintain an optimal velocity that
depends on the driver's headway.
Its motion is given by Equation~\eqref{eqn:ov}.
The optimal velocity function introduced in this equation is represented
by the relationship between the driver's headway and velocity.

The OV model has a simple form with a single parameter of sensitivity
$a$.
This simplicity allows us to understand the mechanisms of phenomena, and
also predict the results of experiments.
The disadvantage of this model, on the other hand, is a little allowance
for control systems to turn up the parameter.
There is indeed a need for collaborations between the research domains
of engineering and physics.
To begin with, however, we restrict our attention to the OV model.
More details are found in~\cite{Bando1995}.

The traffic simulation program based on the OV model is straightforward,
as the model itself has a simple form:

@@

The computation in Step 3 exploits a lot of loop procedures to derive
locations and velocities of agents.
This part is the most compute-intensive, and thus can be accelerated by
the GPU.

@@

GPU performance is dominated by the program design.
GPU-accelerated programs are typically divided into two pieces of code.
The CPU code plays a role of a master thread that controls the program
flow.
The GPU code, on the other hand, spawns a bunch of worker threads to
execute compute-intensive parts of the program in parallel, thus
accelerating the overall program.
What is often argued in performance optimization is how to parallelize
the compute-intensive parts into threads.
This is actually the well-studied problem in the literature.
What is not really understood yet is when to offload the program on to
the GPU.
This paper explores two schemes that use the GPU at different timings to
see how GPU performance is affected.

@@

We focus on the one-dimensional OV model~\cite{Bando1995}, which can be
applied to freeways simulation.
The original program~\cite{OV_Program} of this simulation is written in
C, without any parallelization.
We use the same simulation parameters as the original default setup,
while increasing the number of agents (cars) to see how simulation
performance varies. 
The evaluation is conducted by comparing our GPU implementations with
the original CPU implementation.

@@

So far we have studied a single process of traffic flow.
In reality, however, there are multiple processes of traffic flow.
Real-time feedback from the real road conditions may also be
additionally required to deploy traffic simulation.
All these pieces of simulation must meet the quality-of-service (QoS)
requirement for users.
Now, this turns into a real-time systems problem.
We have developed a lot of solutions to use the GPU in real-time
systems~\cite{Kato2011_3, Kato2011_2, Kato2011_1, Kato2012}.
We believe that some of these work could contribute to GPU-accelerated
real-time traffic simulation.

>>>>

\title{Toward GPU-accelerated Traffic Simulation and Its Real-Time Challenge}

\author {
Manato Hirabayashi, Shinpei Kato, Masato Edahiro\\
\textit{Department of Information Engineering}\\
\textit{Nagoya University}\\
\and
Yuki Sugiyama\\
\textit{Department of Complex Systems Science}\\
\textit{Nagoya University}\\
}

@@

\begin{abstract}
 Traffic simulation is a growing domain of computational physics.
 There are many applications in our life and industry that benefit
 from traffic simulation.
 A core challenge of this science research, however, is posed due to
 an unbounded scale of computation.
 In this paper, we explore an advantage of using the graphics processing
 unit (GPU) to support its computation.
 Two schemes are presented to study how to maximize GPU
 performance in the context of traffic simulation.
 Our experimental results show that even the most basic GPU
 implementation based on our schemes accelerates simulation by about
 five times over the original CPU implementation.
 We also demonstrate that additional orders-of-magnitude
 improvements could be achieved by overcoming the current hardware
 limitation.
\end{abstract}

@@

Transportation systems underlie our life and industry, as part of
societal infrastructure.
Traffic flow is particularly a major factor that influences economy.
According to the Japanese government report, economic losses due to
traffic congestion could reach one hundred billion dollars per year in
Japan, through degradation of transport efficiency, energy consumption,
and environmental poisoning.
Albeit a major issue of transportation systems, the mechanism of traffic
congestion is not well-explained in the literature.
What is particularly important is ``phantom jam'', which often happens
on freeways without any accident.
This phantom jam should be avoided by science, given that it is
a physical phenomenon but not one caused by human errors.
However, we have not been aware of scientific and engineering solutions
to such real-world problems yet.

In physics, traffic flow is described by mathematical
models~\cite{Bando1995, Bando1995_2, Kerner1993, Nagel1992}.
While their mathematical expressions are not identical, they are
commonly compute-intensive forms.
For example, traffic simulation based on the Optimal Velocity (OV)
model~\cite{Bando1995, Bando1995_2} computes locations and velocities of
agents every sampling period, solving OV equations. 
Specifically, the location $x_n$ of the $n$th agent is described by the
following equation, where $\Delta x_n = x_{n+1} - x_n$ is a distance to
a preceding agent, $a$ is a sensitivity, and $V()$ is an optimal
velocity function:
\begin{eqnarray}
 \label{eqn:ov}
 \frac{d^2 x_n}{d t^2} = a \left\{V(\Delta x_n) - \frac{d x_n}{d t}\right\}.
\end{eqnarray}

@@

Computational workload is not the only issue of concern.
Deployment of traffic simulation may also require real-time feedback
from the real world, which turns out to be a cyber-physical systems
(CPS) problem indeed.
Real-time traffic simulation is another challenging issue, where
the rate and the preciseness of simulation must be traded to meet the
requirement of a given scenario.
For instance, simulation of emergent evacuation may want very high-rate
computation, even sacrificing preciseness of simulation to some extent.
Unfortunately, none of those challenges has been explored yet, largely
due to a lack of multidisciplinary collaborations between computationsl
physics and computer science.

@@

The rest of this paper is organized as follows.
Section~\ref{sec:assumption} describes the basic assumption and
terminology used in this paper.
Section~\ref{sec:traffic_simulation} explains an overview of
traffic simulation and its literature.
Section~\ref{sec:gpu_implementations} presents our schemes of GPU
implementations for traffic simulation.
Section~\ref{sec:evaluation} evaluates the advantage of using our
schemes, and provides an insight into future work.
Section~\ref{sec:conclusion} concludes this paper.

@@

While we are intensively interested in CUDA and GPUs, the notion of
accelerating traffic simulation presented in this paper is widely
applicable to heterogeneous architectures, where CPUs and 
accelerators co-exist to complement each other. 
GPUs are currently the most well-recognized forms of accelerators, but
emerging alternatives include the Intel Many Integrated Core (MIC) and
AMD Fusion architectures.
However, their programming models are almost identical in that a master
thread running on the CPU manages the program flow, while parallelized
threads are offloaded on to accelerators to boost computing performance.
CUDA, OpenCL, HMPP, and AMP are all such programming languages that
support heterogeneous architectures under this concept.

@@

The motivation to study traffic flow is to understand the property of
traffic flow, mainly why traffic congestion occurs.
The study of traffic flow has a long history, and many engineering and
physical models exist.
This paper is particularly focused on the OV model~\cite{Bando1995,
Bando1995_2}.
In this model, a driver is supposed to maintain an optimal velocity that
depends on the driver's headway.
Its motion is given by Equation~\eqref{eqn:ov}.
The optimal velocity function introduced in this equation is represented
by the relationship between the driver's headway and velocity.

The OV model has a simple form with a single parameter of sensitivity
$a$.
This simplicity allows us to understand the mechanisms of phenomena, and
also predict the results of experiments.
The disadvantage of this model, on the other hand, is a little allowance
for control systems to turn up the parameter.
There is indeed a need for collaborations between the research domains
of engineering and physics.
To begin with, however, we restrict our attention to the OV model.
More details are found in~\cite{Bando1995, Bando1995_2}.

The traffic simulation program based on the OV model is straightforward,
as the model itself has a simple form.
In this paper, we abstract its algorithm by the following stages:

@@

The computational workload in Step 3 exploits a lot of loop executions
to derive locations and velocities of agents.
This is the most compute-intensive part of the algorithm.
The primary goal of this paper is to find out the parallelism of this
workload that can be applied to the GPU.
By nature, traffic simulation based on the OV model is very scalable in
terms of performance with respect to the number of agents.
The location and velocity of each agent depends on only those of
neighbors.
It is worth exploring a breakthrough of this research domain by
coordinating computational physics and computer science technology.

@@

The primary contribution of this paper is to identify the fundamental
schemes of GPU implementations for traffic simulation.
It is very important to understand that GPU performance is dominated by
the program design.
We first reason about the motivation of our schemes.

GPU-accelerated programs are typically divided into two pieces of code.
The CPU code plays a role of a master thread that controls the program
flow.
The GPU code, on the other hand, spawns a bunch of worker threads to
execute compute-intensive parts of the program in parallel, thus
accelerating the overall program.
What is often argued in performance optimization is how to parallelize
the compute-intensive parts into threads.
This is actually the well-studied problem in the literature.
What is not really understood yet is when to offload the program on to
the GPU.
This paper explores two schemes that use the GPU at different timings to
see how GPU performance is affected.

@@

We focus on the one-dimensional OV model~\cite{Bando1995, Bando1995_2},
which can be applied to freeways simulation.
The original program~\cite{OV_Program} of this simulation is written in
C, without any parallelization.
We use the same simulation parameters as the original default setup,
while increasing the number of agents (cars) to see how simulation
performance varies. 
The evaluation is conducted by comparing our GPU implementations with
the original CPU implementation.

@@

So far we have studied a single process of traffic flow.
In reality, however, there are multiple processes of traffic flow.
Real-time feedback from the real road conditions may also be
additionally required to deploy traffic simulation.
All these pieces of simulation must meet the quality-of-service (QoS)
requirement for users.
Imagine how many potential users exist in the real-world transportation
system.
We are facing a core challenge of next-generation real-time systems.

We have developed basic solutions to use the GPU in real-time
systems~\cite{Kato2011_3, Kato2011_2, Kato2011_1, Kato2012}.
Resource management with scheduling, synchronization, and memory
management of GPU applications has been addressed in our previous
work.
We believe that a grander vision of GPU-accelerated traffic
simulation will need an integration of cutting-edge real-time systems
technology and computational physics.
